---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

# Tracking the Unconscious: Neural evidence for the retention of unaware information in visual working memory

[<img alt="alt_text" src="https://img.shields.io/badge/OSF-https://osf.io/gkmsy/-337AB7"/>](https://osf.io/gkmsy/)

This repository contains the code to reproduce the analysis, figures and tables of the paper *Tracking the Unconscious: Neural evidence for the retention of unaware information in visual working memory* by Gambarota et al. The repository contains also the code to run the experiment using Psychopy along with raw EEG and behavioral data.

## Structure

- The `data` folder contains the raw and cleaned EEG and behavioral data:

```{r}
#| echo: false
fs::dir_tree("data", recurse = TRUE, type = "directory")
```

- The `experiment/` folder contains python scripts to run the experiment.

```{r}
#| echo: false
fs::dir_tree("experiment", recurse = TRUE)
```

- `tables/` and `figures/` contains the tables and figures from the paper created with the `02_analysis-behavioral.R`.
- the `01_pre-processing-behavioral.R` script cleans the raw behavioral data
- the `02_analysis-behavioral.R` script runs the multilevel models for the CDT accuracy and SDT analysis
 creating also tables and figures
- the `03_analysis-cda.R` script runs the analysis on ERP data
- the `brain-vision-analyzer-script.ehtp` is the EEG/ERP pre-processing script

Given the storage limits, the raw EEG data are stored on OSF within the `eeg-raw-data` folder. The content of this folder need be downloaded and placed within the `data/raw/eeg` folder before running the scripts.

All the packages are managed using `renv`. Before running the code open the `cda-subliminal.Rproj` file and run `renv::restore()` to create the custom library with all packages. If the `renv` package is not installed run `install.packages("renv")`.

## Session

```{r}
#| echo: false
sessioninfo::session_info()$platform
```

